hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2024-12-06 inst/manuscript/paper.Rmd
write multiple sections in Package Overview and clean up some style stuff
also render paper as JOSS just to see what it looks like as PDF

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids. Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. `hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. `hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. `hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.

@@

The resulting effective value approach has now been widely applied in salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 were have been applied to various species and populaions [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`. Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for `hatchR`. The first is a R package distributed via CRAN that allows users the most customizable application of these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on `hatchR`'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for change arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum or $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

We demonstrate the utility of the check summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year long dataset with temperature readings every thirty minutes.

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568) 
# create empty dataframe
year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))

@@

# take temps from a random normal dist with mean 10 sd 3 for every date time combo in dates and append to column (temp) in year_sim
year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
  abs()

@@

# it is useful to plot again to check your summarized data
plot_check_temp(data = year_sim_summ,
                dates = date,
                temperature = daily_temp,
                temp_min = 0, # temp_min and max lines are user customizable
                temp_max = 15)
```

@@

```{r, eval=FALSE}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

data_sim <- matrix(NA, 1, 2) |> data.frame()
data_sim[,1] <- temps_sim
data_sim[,2] <- dates_sim

@@

### predict_phenology

#smallmouth bass
smb_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = smb_mod$func)

# catfish
catfish_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = cat_mod$func)

# sturgeon
# note that 16 C is pretty far out of range of temps for model fit, not best practice
sturgeon_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = sturgeon_mod$func)

@@

```{r, echo=FALSE}
# summary for all species
all_res <- data.frame(matrix(NA, 3, 2))
colnames(all_res) <- c("start", "stop")
all_res[1,] <- smb_res$dev.period
all_res[2,] <- catfish_res$dev.period
all_res[3,] <- sturgeon_res$dev.period

all_res <- all_res |> 
  mutate(days = stop-start)
all_res$Species <- c("Smallmouth Bass", "Channel Catfish", "Lake Sturgeon")
```

@@

>>>>

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. `hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. `hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. `hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.

@@

The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`. Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for `hatchR`. The first is a R package distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on `hatchR`'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though the model table includes more complex models developed by @beacham1990) as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

# create empty dataframe
year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))

@@

# take temps from a random normal dist with mean 10 sd 3 
# for every date time combo in dates and append to column (temp) in year_sim
year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
  abs()

@@

# it is useful to plot again to check your summarized data
plot_check_temp(data = year_sim_summ,
                dates = date,
                temperature = daily_temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 15)
```

@@

```{r}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

data_sim <- matrix(NA, 30, 2) |> data.frame()
data_sim[,1] <- temps_sim
data_sim[,2] <- dates_sim

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")

@@

@@

Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.

```{r}
#model fits
smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared

We additionally provide the model estimates for hatch timing for each of the three species used to generate custom models in Figure XXX.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### predict_phenology

#smallmouth bass
smb_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = smb_mod$func)

# catfish
catfish_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = cat_mod$func)

# sturgeon
# note that 16 C is pretty far out of range of temps for model fit, not best practice
sturgeon_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = sturgeon_mod$func)

# summary for all species
all_res <- data.frame(matrix(NA, 3, 2))
colnames(all_res) <- c("start", "stop")

all_res$start <- c(catfish_res$dev.period$start, 
                   smb_res$dev.period$start, 
                   sturgeon_res$dev.period$start)

all_res$stop <- c(catfish_res$dev.period$stop,
                  smb_res$dev.period$stop, 
                  sturgeon_res$dev.period$stop)

all_res <- all_res |> 
  mutate(days = ceiling(stop-start),
         index = c(17,16.5,16))

all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")

ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))
```

@@

We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.
